The Social Conditioning of Fear

### 2023-12-07 scripts/manuscript.qmd
complete

---
title: "The Social Conditioning of Fear"
author:
  - Alejandra Pedraza
  - W. Jamie Yang
abstract: "TBW" # this is where abstract should be pasted 
format: html
wordcount-html: default
toc: true
toc-location: left
toc-title: Table of Contents
highlight-style: pygments
editor: visual
execute:
  eval: true # change to false before final submission
  echo: true
  message: false
  warning: false
fig-width: 6.5
fig-height: 8
code-fold: true
code-summary: "View Code"
embed-resources: true
bibliography: references.bib
link-citations: true
---

@@

Especially, the role of gender in fear of crime has been well documented in victimization research [@cops2011; @snedker2006; @schnitzer2010; @stanko1993]. Early scholarship emphasized that women tend to report more fear than men [@stanko1990], with some initial research investigating the seeming disjuncture between women's fear and their actual experiences of victimization, as women were often found to report high levels of fear for crimes that they were at relatively little risk of experiencing [@hough1983]. Subsequent studies suggested that women's fear should be understood in the context of the many potential threats that women confront, including their heightened risk of sexual assault [@stanko1993]. Therefore, fear at night is gradually perceived as a manifestation of gender-based violence with sexual assault implications, though acknowledging that sexual assault is merely one among many potential forms of crime. However, the more insidious and everyday forms of harassment are frequently not captured by common survey instruments----they regularly privilege the more obvious or extreme forms of victimization and fail to account for the routine forms of violence that women confront [@moran2003]. We focus our attention on this less intrusive but omnipresent form of everyday violence and examine the unequal effects it has on different populations. We ask: what groups of people are more likely to have experienced fear of walking alone at night in their neighborhood?

Historically, despite all the aforementioned contexts and individual-level characteristics being identified as salient to individuals' self-reported fear at night, they were usually studied in isolation. Notably, scholars have tended to focus either on individual-level characteristics, such as race and sexual orientation, or on physically defined contexts like schools and neighborhoods and social contexts like networks, families, with little consideration of the joint contributions of multiple contexts and multilevel effects simultaneously. To fill this gap in knowledge, we hypothesize that, similar to "the social determinants of health" [@krieger2001], there exist the social determinants of fear----intersectional social positionalities impact the likelihood of individuals' sense of threat from social surroundings, i.e., fear at night.

@@

>>>>

---
title: "The Social Conditioning of Fear"
author:
  - Alejandra Pedraza
  - W. Jamie Yang
abstract: "Fear at night has been perceived as a form of gender-based violence with sexual assault implications. However, this more insidious and everyday forms of harassment are frequently not captured by common survey instruments. In this study, we investigate the inequalities in terms of fear of walking alone at night among the US adult population and mapping out the patterns of disparities along the axes of race, gender, sexual orientation, citizenship status, and education. We use the 2022 data from the General Social Survey (GSS) and the data science tools from EDLD 651 class to delineate the data distribution and basic correlations between fear of walking alone at night and our other variables of interest. Our findings suggest that fear at night differs by race, gender, subjective health, and neighborhood diversity. Consistent with previous research, women were also more likely than men to report fear at night. Importantly, groups with multiply marginalized social identities are likely to report higher rates of fear at night." # this is where abstract should be pasted 
format: html
wordcount-html: default
toc: true
toc-location: left
toc-title: Table of Contents
highlight-style: pygments
editor: visual
execute:
  eval: false # change to false before final submission
  echo: true
  message: false
  warning: false
fig-width: 6.5
fig-height: 8
code-fold: true
code-summary: "View Code"
embed-resources: true
bibliography: references.bib
link-citations: true
---

@@

Especially, the role of gender in fear of crime has been well documented in victimization research [@cops2011; @snedker2006; @schnitzer2010; @stanko1993]. Early scholarship emphasized that women tend to report more fear than men [@stanko1990], with some initial research investigating the seeming disjuncture between women's fear and their actual experiences of victimization, as women were often found to report high levels of fear for crimes that they were at relatively little risk of experiencing [@hough1983]. Subsequent studies suggested that women's fear should be understood in the context of the many potential threats that women confront, including their heightened risk of sexual assault [@stanko1993]. Therefore, fear at night is gradually perceived as a manifestation of gender-based violence with sexual assault implications, though acknowledging that sexual assault is merely one among many potential forms of crime. However, the more insidious and everyday forms of harassment are frequently not captured by common survey instruments---they regularly privilege the more obvious or extreme forms of victimization and fail to account for the routine forms of violence that women confront [@moran2003]. We focus our attention on this less intrusive but omnipresent form of everyday violence and examine the unequal effects it has on different populations. We ask: what groups of people are more likely to have experienced fear of walking alone at night in their neighborhood?

Historically, despite all the aforementioned contexts and individual-level characteristics being identified as salient to individuals' self-reported fear at night, they were usually studied in isolation. Notably, scholars have tended to focus either on individual-level characteristics, such as race and sexual orientation, or on physically defined contexts like schools and neighborhoods and social contexts like networks, families, with little consideration of the joint contributions of multiple contexts and multilevel effects simultaneously. To fill this gap in knowledge, we hypothesize that, similar to "the social determinants of health" [@krieger2001], there exist the social determinants of fear---intersectional social positionalities impact the likelihood of individuals' sense of threat from social surroundings, i.e., fear at night.

@@

Previous research on fear at night has attended to differences between women and men, and among sexual minority populations [@meyer2014; otis2007]. This line of research, however, has not fully examined the intersectional effects of various social dimensions on fear at night, such as race, sexual orientation, gender, and community composition. To be specific, although significant gender differences have been documented [@cops2011; @stanko1990], wherein women are more likely than men to report fear, research has failed to consider the complex ways in which other social dimensions such as race may intersect with regard to such fear.

In this project, we used nationally representative survey data to examine racial, sexual, gender, health, and community contextual differences in reports of being afraid to walk alone at night. Our findings suggest that fear at night differs by race, gender, subjective health, and neighborhood diversity. Consistent with previous research, women were also more likely than men to report fear at night. Further analyses revealed that there exists a significant interaction between race and gender in report of fear. Importantly, groups with multiply marginalized social identities are likely to report higher rates of fear at night.

Our assessment of fear relies on a single measure---fear of walking alone at night---that does not explicitly ask respondents about what, in particular, they fear. Potential objects or events in question would be robbery, sexual assault, verbal harassment, or sexual orientation- or religion-based hate crimes. And with common sense, we know that people embodying different identities and self-presentations will be vulnerable to different forms of street crimes. Further research should explore the particular nature of this fear, and whether it is based on crime victimization experience or verbal harassment.

Despite limitations of this study, our findings provide a good understanding of sociodemographic differences in fear at night. We uncovered an important relation between fear and disadvantaged status: racial and ethnic minorities, women, and individuals of better subjective health are more likely to report fear of victimization than their privileged counterparts.

The R tools we learned from the class "EDLD 651: Intro to R" have been extremely helpful in cleaning and visualizing the data.
